---
title: "R Notebook Coursework"
output:
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_depth: 5
---
```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning = FALSE, message = FALSE)
```

```{r}
library("here")
library("tidyverse")
library("lubridate")
library("ggrepel")
library("gapminder")
library("readr")
library("janitor")
library("stringr")
```

```{r}
# importing data
trips <- read_csv(here("data", "rides_2019_2020_extract.csv"))
stations <- read_csv(here("data", "bikeshare_locations.csv"))

# cleaning names
trips <- clean_names(trips)  
stations <- clean_names(stations)

# removing unwanted columns 
trips <- trips %>%
  select(-c(ride_id, is_equity))

stations <- stations %>%
  select(-c(installed, install_date, removal_date, se_anno_cad_data, owner, 
            temporary_install))

# correcting duration 
trips <- trips %>%
  mutate(duration = end_date - start_date) 

# remove any negative duration 
trips <- trips %>%
  filter(duration > 0)


# correcting member types
trips <- trips %>%
  mutate(member_casual = str_replace_all(member_casual, pattern = "casual", 
                                         replacement = "Casual")) %>%
  mutate(member_casual = str_replace_all(member_casual, pattern = "member", 
                                         replacement = "Member")) 

# correcting NAs in latitude, longitude, start_names, end_names
# fill in missing lat, lng
trips <- trips %>%
  left_join(stations, by = c("start_station_name" = "address")) %>%
  rename(start_longitude = longitude, start_latitude = latitude) %>%
  left_join(stations, by = c("end_station_name" = "address")) %>%
  rename(end_longitude = longitude, end_latitude = latitude) %>%
  select(colnames(trips) | contains("longitude") | contains("latitude"))

# fill in missing start and end station names 
trips <- trips %>%
  left_join(
    stations,
    by = c("start_longitude" = "longitude", "start_latitude" = "latitude")) %>%
  rename(start_address = address) %>%
  left_join(
    stations, 
    by = c("end_longitude" = "longitude", "end_latitude" = "latitude")) %>%
  rename(end_address = address) %>%
  select(colnames(trips) | contains("start_address") | contains("end_address"))


# reduce the names columns by merging together the imported and original ones
trips <- trips %>%
  mutate(start_station_name = coalesce(start_station_name, start_address)) %>%
  mutate(end_station_name = coalesce(end_station_name, end_address)) %>%
  select(-c(start_address, end_address))

# reduce the geo points colums by mergin together the imported and original ones 
trips <- trips %>%
  mutate(start_lat = coalesce(start_lat, start_latitude)) %>%
  mutate(start_lng = coalesce(start_lng, start_longitude)) %>%
  mutate(end_lat = coalesce(end_lat, end_latitude)) %>%
  mutate(end_lng = coalesce(end_lng, end_longitude)) %>%
  select(-c(start_longitude, end_longitude, start_latitude, end_latitude))

# use the longitude and latitude coordinates to turn them into distance between 
# start/end points - via the Haversine formula
trips <- trips %>%
  mutate(earth_radius = 6371) %>%
  mutate(start_lat_rad = start_lat * pi/180) %>%
  mutate(end_lat_rad = end_lat * pi/180) %>%
  mutate(change_lat = (end_lat - start_lat) * pi/180) %>%
  mutate(change_lng = (end_lng - start_lng) * pi/180) %>%
  mutate(a = (sin(change_lat / 2)^2 + cos(start_lat_rad) * cos(end_lat_rad) * 
                sin(change_lng / 2)^2)) %>%
  mutate(c = 2 * asin(sqrt(a))) %>%
  mutate(distance_km = c * earth_radius) %>%
  select(colnames(trips) | contains("distance_km"))
```


# **Introduction**
## Motivation for the work below 
  The main aim of this coursework is to provide insights into bike sharing 
  patterns and features from the raw data that has been made available by 
  Capital Bikeshare. 
  With the idea that results gotten at the end of this 
  coursework, can answer interesting questions related to the features of the 
  sample that we have available to us. But also those outcomes can lead 
  to new questions, which require either more detailed data or more 
  sophisticated methods. 
  
  The topic of bike sharing is worth examining, as shared public transport, such 
  as e-cotters and bikes have been becoming more and more popular in recent 
  years. And this trend is likely to continue to do so, as gas prices remain 
  relatively high, tighter emission standards for vehicles are getting passed, 
  and congestion remains a prominent issue, making conventional transport
  uneconomical to use for personal transportation.
  
# **Data**
## Raw Data
  The data used for this report comes from Capital Bikeshare's trips history 
  data which is publicly available on their website 
  https://ride.capitalbikeshare.com/system-data. The raw data includes two 
  datasets. `trips` gives us information about all the rides that have started 
  between the periods of 1st of January 2019 and 31st of December 2020. And 
  `stations` inform us about all the available bike stations across Capital 
  Bikeshare's network.
  
  The raw data files include 16 and 17 variables each. Some of those various
  variables will be utilized, but some won't and therefore will be removed. As 
  they are either not recorded properly or do not give us insightful information. 
  
## Main retained Variables 
  Here is a brief summary of the main retained variables from both datasets:

  + `duration` - duration of a single trip
  + `start_date` and `end_date` -  when has a trip started and ended 
  + `start_station_name` and `end_station_name` - from which bike station a trip
  starts and at which station the trip ends
  + `address` - a street location of a bike station
  + `start_station_id` and `end_station_id` - same as the above, but instead of 
  names, using the unique numbers of each bike station
  + `objectid` - an unique number of an individual bike station
  + `bike_number` - an unique number of the bike being used
  + `member_casual` - a type of user, indicating whether they have purchased a 
  casual "pass" ride plan, or have become a member by getting a "membership" ride 
  plan
  + `rideable_type` - type of bicycle being used 
  + `start_lat`, `start_lng`, `end_lat`, `end_lng` - the starting and ending 
  coordinates of a single trip
  + `latitude`, `longitude` - the coordinates of an individual bike station


## Excluded variables 
  Here is a summary of all the variables that were eliminated from both datasets:
  
  + `ride_id` - unique number of a ride, which was not utilized at any of the 
  steps carried in the report later on, thus removed 
  + `is_equity` - not made clear what it stands for and was not utilized
  at any of the steps carried in the report later on, thus removed 
  + `installed` - whether a station has been installed, but as all were of the 
  status installed, the variable was obsolete and thus removed 
  + `install_date` and `removal_date` - presumably the installation and removal
  dates of single stations, but as for all stations those both variables indicate
  1st January 1970, the variables were obsolete and thus removed 
  + `se_anno_cad_data` - not made clear for what it stand, but as it was filled 
  with only NA's, it was not utilized at any of the steps carried in the report
  later on, thus removed 
  + `owner` - indicates who owns the stations, but it was not utilized at any of
  the steps carried in the report later, thus removed 
  + `temporary_install` - as none of the stations was a temporary install, this 
  variable was obsolete and thus removed 

## Creation of new variables
  A few additional variables were either modified or added, here is a summary of
  those:
  
  + `duration` - was modified by calculating the difference between
  `start_date` and `end_date`, as it previously it was filled with many missing 
  values
  + `member_casual` - was modified by standardizing the format for *Members* and 
  *Casual* users
  + `start_lat`, `start_lng`, `end_lat`, `end_lng` - were modified, by in filling 
  their missing values, via the dataset `stations`
  + `distance_km` - a new variable, that is based on the Haversine formula,
  calculating the distance in kilometers between the start and end points of a 
  journey, based on `start_lat`, `start_lng`, `end_lat`, `end_lng`


# **Questions**
## Question 1 
  **How has the demand for shared bike riding provided by Capital Bikeshare changed
  between 2019 and 2020 on a monthly basis?**
  
  + Operationalization: The demand for shared bike riding will be measured by the 
  total number of rides that have occurred during a particular period of time. In 
  the case of *Question 1*, the time period will be each month of either 2019 
  or 2020. 
  
## Question 2
  **During the day, are *Casual* users using the services provided by Capital Bikeshare to a greater extent than *Member* users?**
  
  + Operationalization: The utilization variable will be measured by the hours 
  spent on journeys. As this is the only measurement that we have to base the 
  actual use of the service. Further, the journey times will be ordered, 
  depending on which part of the day they have started. As well as by what type 
  of user they have been carried out.

## Question 3
  **People that opt for an electric bike, can either leave it at a bike station free of charge or leave it at any public bike rack for an additional 2$ charge.**
  **Is this choice related to the distance between a journey's start and end points or the duration of it? And has this behavior been consistent over time?**
  
  + Operationalization: To differentiate between which users have been 
  charged an additional 2$, we will look at which journeys made via an electric 
  bike that have no end station, but rather just end coordinates. Based on these 
  findings, we will examine, if this decision has a relation with the 
  duration of the ride in minutes and the distance in kilometers between the 
  start and end points of the journey. It is important to note that we don't 
  imply any causality. 
  
## Question 4
  **Is the distance traveled between the start and end bike stations, indicative of the riding plan chosen by a user?**

  + Operationalization: The distance between the start and end bike stations will
  be measured in kilometers. And further, those distances will be grouped depending 
  on the ride plan that users have opted to choose. Measuring the density that 
  each user group represents, based on their distances.
  
## Question 5 
  **Which are the 5 most popular journeys among Capital Bikashare's users and how long does it take them on average to complete it?**
  
  + Operationalization:  The journey variable will be consisting of a combination
  between the start and end station of a bike ride. The 5 most popular journeys
  are going to be chosen based on their frequency. While the average times of 
  completion are going to be measured in hours by a median. With further 
  information about the spread, provided by an interquartile range.

# **Analysis**
## Analysis to Question 1 

```{r}
# Question 1 
# setting the graph 
tripsQ1 <- trips %>%
  mutate(start_year = format(trips$start_date, format = "%Y")) %>%
  mutate(start_month = format(trips$start_date, format = "%b")) %>%
  mutate(start_month = factor(start_month, levels = month.abb))

q1_plot <- ggplot(
    data = tripsQ1,
    mapping = aes(x = start_month))

q1_graph_backdrop <- geom_bar(
    data = transform(tripsQ1, start_year = NULL),
    fill = "grey85")

q1_scaling <- scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.05)))

q1_labels <- labs(
    title = "Demand for Capital Bikeshare's service throughout 2019 and 2020",
    x = "Month",
    y = "Amount of rides taken")

q1_theme_settings <- theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey80"),
    panel.grid.minor.y = element_line(colour = "grey80"),
    strip.text.x = element_text(size = 13))

# building the graph 
  q1_plot + 
  q1_graph_backdrop +
  geom_bar () +
  facet_wrap(facets = vars(start_year)) +
  q1_scaling +
  q1_labels +
  q1_theme_settings
```

 What can be immediately seen from the "Demand for Capital Bikeshare's service 
 throughout 2019 and 2020" figure is that 2020 has seen far fewer rides taken, 
 relative to 2019. 
 It is interesting to notice that there is a sharp drop in April 2020, after 
 which the demand for the service is not able to catch up to the 2019 figures. 
 This perhaps may be due to COVID-19 measures that have been taking place at 
 that time.
 
 Further, it can be noticed, that the months of November, December, January, and 
 February see far fewer rides being taken, relative to the rest. This pattern
 holds true for both years in mind. The decline may be due to less favorable
 weather that there might be for riding a bike.
 
## Analysis to Question 2 

```{r}
# Question 2 
# setting the graph 
tripsQ2 <- trips %>%
  mutate(duration = as.double(duration / 3600)) %>%
  mutate(start_hour = format(trips$start_date, format = "%H"))

q2_plot <- ggplot(
    data = tripsQ2,
    mapping = aes(x = start_hour, y = duration, colour = member_casual, 
                  fill = member_casual))

q2_scaling <- scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.14)),
    n.breaks = 7) 
  
q2_colour_scaling <- scale_colour_manual(values = c("#990000", "#1380A1"))

q2_theme_settings <- theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey80"),
    panel.grid.minor.y = element_line(colour = "grey80"))
    # axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 

q2_labels <- labs(
    title = "Usage of Capital Bikeshare's service",
    x = "Time of day",
    y = "Total usage hours",
    fill = "User type",
    colour = "User type") 

# building the graph
q2_plot +
  geom_col() +
  q2_scaling +
  q2_colour_scaling +
  q2_labels +
  q2_theme_settings
```

The graph "Usage of Capital Bikeshare's service" shows that in every hour of the 
day, *Casual* users tend to spend more time on the bike sharing platform, 
relative to  *Members* users.
Further, it is interesting to notice that in the morning between 7 AM and 8 AM 
there is a sudden surge, in the duration of journeys being made, regardless of the
type of user. However, it is not until in the afternoon after 3 PM that the most
time is accumulated of using the service. Those times tend to be dominated by 
*Casual* members.

For both user types, the most popular time for starting a journey is 5 PM, 
whilst the least popular is 4 AM.

## Analysis to Question 3 

```{r}
# Q3
# setting up the graph
trips_Q3 <- trips %>%
  mutate(start_date1 = as.Date(start_date)) %>%
  filter(rideable_type == "electric_bike") %>% 
  filter(distance_km < 47) %>%
  mutate(duration = as.double(duration / 60)) %>%
  mutate(end_station_returned = as.logical(is.na(end_station_name))) %>%
  mutate(end_station_returned = case_when(
    end_station_returned == TRUE ~ "Yes", #not returned to a station
    end_station_returned == FALSE ~ "No")) #returned to a station

q3_plot <- ggplot(
    data = trips_Q3,
    mapping = aes(x = start_date1, y = distance_km, colour = end_station_returned, size = duration))

q3_colour_scaling <- scale_colour_manual(values = c("#1380A1", "#990000"))

q3_x_scaling <- scale_x_date(
    limits = as_date(c("2020-06-17", "2020-31-12")),
    date_breaks = "1 month",
    date_labels = "%b")

q3_y_scaling <- scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.3)),
    n.breaks = 6)

q3_labels <- labs(
    title = "Additional charge for e-bikes relation to duration and distance",
    x = "Month (2020)",
    y = "Distance (km)",
    colour = "Additionally charged",
    size = "Duration of journey (minutes)")

q3_theme_formatting <- theme_minimal() +
  theme(
    panel.grid.major.x = element_line(colour = "grey80"),
    panel.grid.major.y = element_line(colour = "grey80"),
    panel.grid.minor.y = element_line(colour = "grey80")) 

# building the graph
q3_plot +
  geom_point(alpha = 0.4) +
  q3_colour_scaling +
  q3_x_scaling +
  q3_y_scaling +
  coord_cartesian(ylim = c(0, 20)) +
  q3_labels +
  q3_theme_formatting
```

Something interesting about the "Additional charge for e-bikes relation to duration 
and distance" figure is that the *Month (2020)* scale starts from mid-June. 
This is the case as e-bikes have been introduced on the 17th of June 2020 and 
simply previous data does not exist. 

The figure makes it immediately clear that the majority of users that use e-bikes 
are opting not to get charged an additional 2$ and return their e-bike to an end 
station. However it can be seen that between September and November, there
has been some increase in the users that have preferred to pay the additional 
charge. Further, interestingly those members have traveled for at least 
more 1 kilometer between their initial and final point.

Moving up the y-axis, it can be seen that for users that have traveled around or 
more than 10km there is an increase in the choices to pay the additional charge. 
Further, it can be noticed that people that have preferred to lock their e-bikes
to a public stand have a much higher journey duration, than those who have returned 
theirs at a Capital Bikeshare station.

## Analysis to Question 4 

```{r}
# Q4
# setting the graph
trips_Q4 <- trips %>%
  mutate(duration = as.double(duration / 60)) %>%
  filter(duration < 120) %>%
  filter(distance_km < 47)

q4_plot <- ggplot(
    data = trips_Q4,
    mapping = aes(x = distance_km, colour = member_casual, fill = member_casual)
  )

q4_colour_scaling <- scale_colour_manual(values = c("#990000", "#1380A1"))
  
q4_x_scaling <- scale_x_continuous(breaks = seq(0, 20, 5))

q4_y_scaling <- scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.1)))
  
q4_text_annotation <- annotate(
    geom = "text",
    x = 15,
    y = 0.5,
    label = "start and end points \nof those journeys \nare identical", 
    size = 4.2,
    stroke = 1.1)
  
q4_top_curve <- annotate(
    geom = "curve",
    x = 11.2,
    y = 0.50,
    xend = 0.15,
    yend = 0.6,
    curvature = -0.1,
    arrow = arrow(length = unit(2.5, "mm")),
    alpha = 0.8)
  
q4_bottom_curve <- annotate(
    geom = "curve",
    x = 11.6,
    y = 0.48,
    xend = 0.15,
    yend = 0.2,
    curvature = 0.15,
    arrow = arrow(length = unit(2.5, "mm")),
    alpha = 0.8)
   
q4_labels <- labs(
    title = "Distance traveled between start and end stations relation to ride plan",
    x = "Distance (km)",
    y = "Density",
    fill = "User type",
    colour = "User type")
  
q4_theme_formatting <- theme_minimal() +
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey80"),
    panel.grid.minor.y = element_line(colour = "grey80")) 

# building the graph
q4_plot +
  geom_density(
    size = 0.9, 
    alpha = 0.3) +
  q4_colour_scaling +
  q4_x_scaling +
  q4_y_scaling +
  q4_text_annotation +
  q4_top_curve +
  q4_bottom_curve +
  q4_labels +
  q4_theme_formatting
```

The "Distance traveled between start and end stations relation to ride plan", 
shows us that for both user types a popular choice is to return their rented 
bike to the same station from which they have taken it initially. 
This is indicated by the both the red and blue line surges along the y axis at 
the 0 km mark.

Another interesting trend is there is a greater chance a user from *Members* to 
prefer returning their bike to a different station than their initial one, 
relative to a *Casual* user. As seen by the area difference of the blue and red
peaks between the 0 and 2 kilometers mark roughly.

And finally this graph shows is that the very little chance of the population 
of users regardless of the type that are willing to make a journey between 
stations that are more than 5 km apart.

## Analysis to Question 5 

```{r}
# Q5 
# setting the graph
popularity <- trips %>%
  drop_na(start_station_name) %>%
  drop_na(end_station_name) %>%
  unite(
    col = "journey",
    c(start_station_name, end_station_name),
    sep = " to "
  ) %>%
  group_by(journey) %>%
  summarise(
    journey_pop = n()
  ) %>%
  top_n(5) %>%
  arrange(desc(journey_pop))

journeys <- trips %>%
  mutate(duration = as.double(duration)) %>%
  drop_na(start_station_name) %>%
  drop_na(end_station_name) %>%
  unite(
    col = "journey",
    c(start_station_name, end_station_name),
    sep = " to "
  )

trips_Q5 <- popularity %>%
  left_join(
    journeys,
    by = "journey"
    ) %>%
  mutate(duration = as.double(duration / 3600)) %>%
  mutate(year = year(start_date)) 

  q5_plot <- ggplot(
    data = trips_Q5,
    mapping = aes(x = duration, y = journey, fill = journey_pop))

  q5_x_scalling <- coord_cartesian(xlim = c(0, 4))
  
  q5_y_scalling <- scale_y_discrete(labels = function(y) str_wrap(y, width = 27))
  
  q5_labels <- labs(
    title = "Rent times for the 5 most popular routes",
    x = "Duration (hours)",
    y = "",
    fill = "Times journey being taken")
  
  q5_theme_formatting <- theme_minimal() +   
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey80"),
    panel.grid.minor.x = element_line(colour = "grey80"),
    legend.position = "top")
  
  # building the graph 
  q5_plot +
    geom_boxplot() +
    scale_fill_viridis_b() +
    q5_x_scalling +
    q5_y_scalling +
    q5_labels +
    q5_theme_formatting
```

From the figure "Rent times for the 5 most popular routes" it can
be seen that all of the 5 most popular journeys, commence and finish at the same 
bike station. Given the results from Question 4, this does not seem surprising.
 
Further, it can be seen that half of the users carrying out those journeys, 
rent out a bike from anywhere 30 minutes to an hour and fifteen minutes. The 
only exception is the fifth most popular journey, which starts and finishes at 
Gravelly Point. Fifty percent of the population done it, tend to 
complete it for anywhere between 20 to 50 minutes.


# **Conclusion** 
## Summary of results 
  The results that we have obtained have given us useful insights concerning
  bike sharing. 
  
  First of all, we found that the overall demand for the Capital Bikeshare in Washington 
  DC has fallen in 2020, in comparison to 2019. Further, we saw that the majority
  of users in terms of time spent using the service are of the *Casual* type.
  With the third question, we saw that a large proportion of e-bike users,
  prefer not to pay 2$ for the convenience of leaving their bike pretty much 
  anywhere. And we examined that there is a relationship between electric 
  bicycle users that tend to opt for the additional charge and substantial distance 
  over a great duration period. With our last two questions, we were able to find 
  out that the majority of journeys, tend to start and end at the same station. 
  For the 5 most popular such trips, users on average rent a bike for 
  just under an hour.
  
## Further questions
  Given the analysis done above, it will be interesting to see how the 
  duration of the rides has differed with actual traveled distance and 
  not only distance between the initial and finishing bike stations. As this will 
  give us more insight into the potential wear and tear of the bikes. But also 
  could be used for predicting average times for various common travel routes.

